# EDS-214 Final Project

### Author
 Amapola Garcia

## Background
This Github repository contains the raw data and R code that is associated to recreate the [Figure 3](Schaefer DouglasA, McDowell WH, Scatena FN, Asbury CE. Effects of hurricane disturbance on stream water concentrations and fluxes in eight tropical forest watersheds of the Luquillo Experimental Forest, Puerto Rico. Journal of Tropical Ecology. 2000;16(2):189-207. doi:10.1017/S0266467400001358) from Schaefer et al. (2000). The repository showcases the pre and post-effect disturbances from Hurricane Hugo 1989. What they found was deposits of some nutrients such as, potassium, increased greatly post-disturbances and stayed elevated for years afterwards.

## Data
In the first inital step we are downloading the four raw datasets that are provided by the Enviromental Data.

 McDowell, W. and International Institute of Tropical Forestry(IITF), USDA Forest Service.. 2024. Chemistry of stream water from the Luquillo Mountains ver 4923064. Environmental Data Initiative.https://doi.org/10.6073/pasta/f31349bebdc304f758718f4798d25458 (Accessed 2026-08-25).

```{r}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
cleaned_data <- read_csv("../output/1_clean_data.csv")
```

## Methods 
The analysis calculates the 9-weeks of moving averages from the chemistry measurments and plots these averages over a time period from 1988 to 1995.


## Results
 The figure down below is a replica representation of figure(3) of Scaefer et al.(2000).
 In the image down below we can see a time series plotted replication that has the Sample_date on the x-axis and the concentration on the y-axis.

After the hurricane, you can see that the concentration levels from Hurrican Hugo increased greatly and stayed elevated throughout the years afterwards. Each of the ion levels are plotted on the y-axis and are found moving aross the average of the 9-week window.

```{r}
#| echo: false
#| message: false
#| warning: false

ggplot(
  cleaned_data,
  mapping = aes(
    x = Sample_Date,
    y = Concentration,
    linetype = site
  )
) +
  geom_line() +
  facet_wrap(
    ~Ions,
    scales = "free",
    ncol = 1,
    strip.position = "left"
  ) +
  labs(
    title = "Concentrations from Hurricane Hugo in Bisley, Puerto Rico"
  )
```
Figure 1: A reproduced version of the Concentrations in Bisley, Puerto Rico streams before and after Hurricane Hugo.


## Reference
Schaefer, Douglas. A., William H. McDowell, Fredrick N. Scatena, and Clyde E. Asbury. 2000. “Effects of Hurricane Disturbance on Stream Water Concentrations and Fluxes in Eight Tropical Forest Watersheds of the Luquillo Experimental Forest, Puerto Rico.” Journal of Tropical Ecology 16 (2): 189–207. https://doi.org/10.1017/s0266467400001358